The corona of a fully convective star with a near-polar flare

### 2023-09-19 src/tex/ms.tex
[showyourwork] overleaf sync

Fig.~\ref{fig:lightcurves} shows the time series using the combined PN and MOS time series together with the OM light curve. We extracted the events time series from EPIC using XMM SAS version 20~\footnote{"Users Guide to the \textit{XMM-Newton} Science Analysis System", Issue 18.0, 2023 (ESA: \textit{XMM-Newton} SOC)} using the \texttt{evselect} task. We selected a circular source region with 20 arcsec radius on all detectors, and a circular source-free background region 120 and 90 arcsec in MOS and PN, respectively. We used the \texttt{epiclccorr} task in XMM SAS to correct for energy and time dependent loss of events\footnote{\url{https://heasarc.gsfc.nasa.gov/docs/xmm/sas/help/epiclccorr/node3.html}, accessed on July 3, 2023}, and extract a soft X-ray time series with a time binning of $200\,$s.

@@

\begin{figure*}[ht!]
    \script{FIGURE_lightcurves.py}
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{figures/lightcurves.png}
        \caption{
         Top panel: Optical Monitoring (OM) light curve. Bottom panel: X-ray (PN, MOS1 and MOS2 combined) light curve. The grey shaded portion defines the flare-only subset of the observations (see Section~\ref{sec:methods:epic} and Table~\ref{tab:specfit}).
        }
        \label{fig:lightcurves}
    \end{centering}
\end{figure*}

@@

 \begin{table}
% \movetableright=-20mm
\centering
    \script{TABLE_specfits.py}
    \caption{XSPEC fits to EPIC spectra for different subsets of observations. Fluxes and luminosities are given in the $0.2-2\,$keV band.}
    \input{output/mcmc_specfit.tex}
        \label{tab:specfit}
\end{table}

@@

We find that the quiescent corona of TIC 277 is best described by two components, similar to many other active M dwarfs~\citep{wargelin2008xray, robrade2010quiescent, behr2023muscles, magaudda2022firsta}. The hotter component represents thermal bremsstrahlung radiation from hot plasma that evaporates from the chromosphere into the coronal flaring loops~\citep{benz2016flare}, which is supported by the observed delay between the optical and the X-ray flare~\citep[Fig.~\ref{fig:lightcurves}, ][]{hawley2003multiwavelength}. The optical flare originates from the lower chromosphere, which is heated by the precipitating particles. This heating then causes evaporation of particles into the flare loop. %A third, very hot, possibly $>40\,$MK component could only be poorly constrained due to a lack of sensitivity to flux above $5\,$keV, and its short lifetime during the peak of the flare.

The range of 6.5-11 MK for the emission measure weighted temperature of TIC 277 places it squarely in the saturated regime with other mid M dwarfs down to spectral type M6~\citep{wright2018stellar, magaudda2020relation, stelzer2022first,robrade2005xray,raassen2003xray,paudel2021simultaneous, foster2020corona}. Flares on the slowly rotating Proxima Cen (M5.5, $P_{\rm rot} = 83\,$d,~\citealt{anglada-escude2016terrestrial}), appear with a range of temperatures at 2-3 MK and 10-20 MK~\citep{gudel2004flares, fuhrmeister2011multiwavelength, fuhrmeister2022high, howard2022mouse} placing it in the transition region between saturated and unsaturated regimes. Only at rotation periods beyond 100 days, flaring activity declines to yield coronal temperatures $<2\,$MK~\citep{wright2018stellar, foster2020corona}.

@@

Recent observations of TRAPPIST-1 b~\citep{greene2023thermal, ih2023constraining} with the James Webb Space Telescope find the planet mostly stripped of an atmosphere, which the authors attribute to the intense flare driven X-ray radiation and energetic particle exposure at the short orbital periods of the innermost planet. For the habitable zone planets, TRAPPIST-1 e and f, simulations suggest that flares alter their chemical composition, in particular ozone, nitric oxide and hydroxide, but that the effects would be below the noise floor of JWST~\citep{chen2021persistence}. However, given the uncertain age of TRAPPIST-1~\citep{burgasser2017age, gonzales2019reanalysis, birky2021improved}, the temperate planets e and f may have been entirely stripped of their atmospheres in the past gigayear(s)~\citep{garraffo2017threatening}.

TIC~277's rapid rotation and UV flux is indicative of a younger age than TRAPPIST-1, although it has not been attributed to any star cluster so far~\citep{schneider2018hazmat}. Interpreting the upcoming observations of TRAPPIST-1 e with JWST will most likely involve its activity history. Studies of its younger counterparts, late M dwarfs like TIC~277, are required to empirically constrain cumulative effects of atmospheric forcing of planets. If its energetic flares commonly occur at high latitudes~\citep{ilin2021giant}, future models of energetic particle exposure of habitable zone planets may have to include age dependent latitudes of particle eruption to reproduce observations.

@@

\begin{table}[]
    \centering
    \caption{Late M dwarfs with known rotation period and X-ray luminosity.}
    \begin{tabular}{lccc}
        ID & SpT &$P_{\rm rot}$ [d] & $\log_{10} \frac{L_X}{L_{\rm bol}} $\\\hline
        TIC 277 (this work) & M7 & 0.19 & $-4$ \\
        TRAPPIST-1 & M8 & 3.3 & $-4$ [1]\\
        LHS 248 & M6.5 & 0.46 & $-3.9$ [2] \\
        NLTT 33370 AB & M7 & 0.16 & $-3.5$ [3] \\
        LP 412-31 & M8 & 0.61 & $-3.1$ [4]\\\hline
    \end{tabular}
    \flushleft
     \footnotesize
    [1] \citet{brown2023coronal}, 
    [2] \citet{cook2014trends}, 
    [3] \citet{williams2015simultaneous}, 
    [4] \citet{stelzer2006simultaneous}    
    %3.3 d from \citep{luger2017sevenplanet}
    \label{tab:latmlxlbol}
\end{table}

>>>>

Fig.~\ref{fig:lightcurves} shows the time series using the combined PN and MOS time series together with the OM light curve. We extracted the events time series from EPIC using XMM SAS version 20\footnote{"Users Guide to the \textit{XMM-Newton} Science Analysis System", Issue 18.0, 2023 (ESA: \textit{XMM-Newton} SOC)} using the \texttt{evselect} task in the full $0-10\,$keV band. We selected a circular source region with 20 arcsec radius on all detectors, and a circular source-free background region 120 and 90 arcsec in MOS and PN, respectively. We used the \texttt{epiclccorr} task in XMM SAS to correct for energy and time dependent loss of events\footnote{\url{https://heasarc.gsfc.nasa.gov/docs/xmm/sas/help/epiclccorr/node3.html}, accessed on July 3, 2023}, and extract a soft X-ray time series with a time binning of $200\,$s.

@@

\begin{figure*}[ht!]
    \script{FIGURE_lightcurves.py}
    \begin{centering}
        \includegraphics[width=0.75\linewidth]{figures/lightcurves.png}
        \caption{
         Top panel: Optical Monitoring (OM) light curve. Bottom panel: Background-subtracted X-ray light curve using the flux in the entire $0-10\,$keV range  (PN, MOS1 and MOS2 combined). The grey shaded portion defines the flare-only subset of the observations (see Section~\ref{sec:methods:epic} and Table~\ref{tab:specfit}).
        }
        \label{fig:lightcurves}
    \end{centering}
\end{figure*}

@@

 \begin{table}
% \movetableright=-20mm
\footnotesize
\centering
    \script{TABLE_specfits.py}
    \caption{XSPEC fits to EPIC spectra for different subsets of observations. Fluxes and luminosities are given in the $0.2-2\,$keV band.}
    \input{output/mcmc_specfit.tex}
        \label{tab:specfit}
\end{table}

@@

We find that the quiescent corona of TIC 277 is best described by two components, similar to many other active M dwarfs, in line with the standard picture for stellar flares~\citep{wargelin2008xray, robrade2010quiescent, behr2023muscles, magaudda2022firsta}. Both the cooler and the hotter component represent thermal radiation from hot plasma that evaporates from the chromosphere into the corona. The hotter component is driven by particles precipitating downward during flares that then heat the chromosphere to high temperatures, and cause additional evaporation into the coronal flaring loops~\citep{benz2016flare}. In our osbervation, this interpretation is supported by the observed delay between the optical and the X-ray flare~\citep[Fig.~\ref{fig:lightcurves}, ][]{hawley2003multiwavelength}. The optical flare originates from the lower chromosphere, which is heated by the precipitating particles. The heating then causes evaporation of particles into the flare loop. In the cooler component, the source of the more quiescent coronal heating is an open question, even for the Sun, where the respective contributions of reconnection and magnetohydrodynamical waves are under investigation~\citep{vandoorsselaere2020coronal}.  %A third, very hot, possibly $>40\,$MK component could only be poorly constrained due to a lack of sensitivity to flux above $5\,$keV, and its short lifetime during the peak of the flare.

The range of 6.5-11 MK for the emission measure weighted temperature of TIC 277 places it squarely in the saturated regime with other mid M dwarfs down to spectral type M6~\citep{wright2018stellar, magaudda2020relation, stelzer2022first,robrade2005xray,raassen2003xray,paudel2021simultaneous, foster2020corona}. Flares on the slowly rotating Proxima Cen (M5.5, $P_{\rm rot} = 83\,$d,~\citealt{anglada-escude2016terrestrial}), appear with a range of temperatures at 2-3 MK and 10-20 MK~\citep{gudel2004flares, fuhrmeister2011multiwavelength, fuhrmeister2022high, howard2022mouse} placing it in the transition region between saturated and unsaturated regimes. Only at rotation periods beyond 100 days, flaring activity declines to yield coronal temperatures $<2\,$MK in M3-M6 dwarfs~\citep{wright2018stellar, foster2020corona}.

@@

%Recent observations of TRAPPIST-1 b~\citep{greene2023thermal, ih2023constraining} with the James Webb Space Telescope find the planet mostly stripped of an atmosphere, which the authors attribute to the intense flare driven X-ray radiation and energetic particle exposure at the short orbital periods of the innermost planet. For the habitable zone planets, TRAPPIST-1 e and f, simulations suggest that flares alter their chemical composition, in particular ozone, nitric oxide and hydroxide, but that the effects would be below the noise floor of JWST~\citep{chen2021persistence}. However, given the uncertain age of TRAPPIST-1~\citep{burgasser2017age, gonzales2019reanalysis, birky2021improved}, the temperate planets e and f may have been entirely stripped of their atmospheres in the past gigayear(s)~\citep{garraffo2017threatening}.

TIC~277's rapid rotation and UV flux is indicative of a younger age than TRAPPIST-1, although it has not been attributed to any star cluster so far~\citep{schneider2018hazmat}. Interpreting the upcoming observations of TRAPPIST-1 e, one of the rocky planets in the habitable zone of the system, with the James Webb Space Telescope will most likely involve its activity history. Studies of its younger counterparts, late M dwarfs like TIC~277, are required to empirically constrain cumulative effects of atmospheric forcing of planets. If its energetic flares commonly occur at high latitudes~\citep{ilin2021giant}, future models of energetic particle exposure of habitable zone planets may have to include age dependent latitudes of particle eruption to reproduce observations.

@@

\begin{table*}[]
    \centering
    \caption{Late M dwarfs with known rotation period and X-ray luminosity.}
    \begin{tabular}{lcccc}
        ID & SpT &$P_{\rm rot}$ [d] & $\log_{10} \frac{L_X}{L_{\rm bol}} $ & energy range [keV]\\\hline
        TIC 277 (this work) & M7 & 0.19 & $-4$ & $0.2-2$ \\
        TRAPPIST-1 [1] & M8 & 3.3 & $-4$ & $0.3-10$\\
        LHS 248 [2] & M6.5 & 0.46 & $-3.9$ & $0.2-2$\\
        NLTT 33370 AB [3] & M7 & 0.16 & $-3.5$ & $0.2-2$ \\
        LP 412-31  [4] & M8 & 0.61 & $-3.1$& \textcolor{red}{soft X-ray band?}\\\hline
    \end{tabular}
     \footnotesize
     \vspace{0.1cm}\newline
    [1] \citet{brown2023coronal}, 
    [2] \citet{cook2014trends}, 
    [3] \citet{williams2015simultaneous}, 
    [4] \citet{stelzer2006simultaneous}    
    %3.3 d from \citep{luger2017sevenplanet}
    \label{tab:latmlxlbol}
\end{table*}
